---
title: "I have the best titles"
author: Amanda Klisavage
subtitle: This change is a test.
---

# Introduction
My really awesome and complete introduction will go here.  

# Materials and methods

Narrative and most code will go here.  Describe what you are doing and show how to do it (with code).

You can do bullets like this:

* The first most important thing
* The second most important thing
* The third most important thing

You can do numbers like this:

1. The first most important thing
2. The second most important thing
3. The third most important thing

See [http://rmarkdown.rstudio.com/](http://rmarkdown.rstudio.com/) for all the amazing things you can do.


Here's my first code chunk.
```{r}
1+2
```

Load any required packages in a code chunk (you may need to install some packages):

```{r, message=F, warning=F}
library(raster)
library(rasterVis)
library(shar)
library(sp)
library(spData)
library(tidyverse)
library(sf)
library(spocc)
library(rbison)
library(maptools)
library(rgdal)
library(knitr)
library(mapview)
library(leaflet)
library(broom)
library(leaflet.extras)
library(shiny)
library(shinydashboard)
knitr::opts_chunk$set(cache=TRUE)  # cache the results for quick compiling
```

## Download and clean all required data
```{r}
############Calling occurrence data and converting into data frame###########
SLF<-occ("Lycorma delicatula", has_coords=TRUE, limit=1000000,)
SLFdf<-occ2df(SLF)

EAB<-occ("Agrilus planipennis", has_coords=TRUE, limit=1000000)
EABdf<-occ2df(EAB)

HWA<-occ("Adelges tsugae", has_coords=TRUE, limit=1000000)
HWAdf<-occ2df(HWA)

ALB<-occ("Anoplophora glabripennis", has_coords=TRUE, limit=1000000)
ALBdf<-occ2df(ALB)

SWW<-occ("Sirex noctilio", has_coords=TRUE, limit=1000000)
SWWdf<-occ2df(SWW)
```

```{r, results='asis'}
data %>% 
  slice(1:10) %>% #show only 1:n rows
  kable(digits=2,align="c")%>% #make table and round to two digits
  kable_styling(bootstrap_options = 
                  c("striped", "hover", "condensed", "responsive")) #apply other formatting
```

Add any additional processing steps here.

# Results

Show tables, plots, etc. and describe them.

```{r, fig.width=6, fig.height=3, fig.cap="Map of completely random data"}

map<-leaflet()%>%
  addTiles()%>%
  addCircleMarkers(lng=SLFdf$longitude, lat=SLFdf$latitude, color="green", popup=SLFdf$date, group="Spotted Lanternfly", fillOpacity=0.2, radius=2)%>%
  addCircleMarkers(lng=EABdf$longitude, lat=EABdf$latitude, color="blue", popup=EABdf$date, group="Emerald Ash Borer", fillOpacity=0.2, radius=2)%>%
  addCircleMarkers(lng=HWAdf$longitude, lat=HWAdf$latitude, color="yellow", popup=HWAdf$date, group="Hemlock Wooly Adelgid", fillOpacity=0.2, radius=2)%>%
  addCircleMarkers(lng=ALBdf$longitude, lat=ALBdf$latitude, color="orange", popup=ALBdf$date, group="Asian Long-horned Beetle", fillOpacity=0.2, radius=2)%>%
  addCircleMarkers(lng=SWWdf$longitude, lat=SWWdf$latitude, color="red", popup=SWWdf$date, group="Sirex Woodwasp", fillOpacity=0.2, radius=2)%>%
  addLayersControl(overlayGroups = c("Spotted Lanternfly", "Emerald Ash Borer", "Hemlock Wooly Adelgid", "Asian Long-horned Beetle", "Sirex Woodwasp"))


```


```{r}
data %>% 
  ggplot(aes(x=x,y=y,col=category))+
  geom_point()
```

# Conclusions

What have you learned?  Are there any broader implications?

# References
